---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r, echo = TRUE}
activityDF <- read.csv("C:/Users/HP/Documents/Coursera/RepData_PeerAssessment1/activity/activity.csv", stringsAsFactor = FALSE)
str(activityDF)
activityDF$date <- as.Date(activityDF$date)
```

## What is mean total number of steps taken per day?

The total number of steps taken per day

```{r, echo = TRUE}
TotalStepsPerDay <- tapply(activityDF$steps, as.factor(activityDF$date), sum)
```



Histogram of total number of steps taken each day

```{r, echo = TRUE}
hist(TotalStepsPerDay, xlim = c(0,22000), ylim = c(0,10), breaks = 20, xlab = "Total Steps Per Day", main = "Histogram of Total Steps Per Day")
```


Mean of the total number of steps taken per day

```{r, echo = TRUE}
mean(TotalStepsPerDay, na.rm = TRUE)
```


Median of the total number of steps taken per day

```{r, echo = TRUE}
median(TotalStepsPerDay, na.rm = TRUE)
```



## What is the average daily activity pattern?

Calculating the average steps taken per interval across all days

```{r, echo = TRUE}
averageStepsPerInterval <- tapply(activityDF$steps, activityDF$interval, mean, na.rm = TRUE)
```


Converting the array into a data frame

```{r, echo = TRUE}
averageStepsPerInterval <- as.data.frame.table(averageStepsPerInterval)
colnames(averageStepsPerInterval) <- c("Interval", "AverageSteps")
head(averageStepsPerInterval)
```


Time-series plot of the 5-minute interval and average number of steps taken, averaged across all days

```{r, fig.width = 12, echo = TRUE}
plot(averageStepsPerInterval$Interval, averageStepsPerInterval$AverageSteps, type = "l")
```


Calculating the 5-minute interval which contains the maximum number of steps, averaged across all days

```{r, echo = TRUE}
averageStepsPerInterval[which.max(averageStepsPerInterval$AverageSteps),]
```


```{r, echo = TRUE}
maxInterval <- averageStepsPerInterval[104, 1]
maxSteps <- averageStepsPerInterval[104,2]
```

The 5-minute interval `r maxInterval` contains the maximum number of steps, `r maxSteps`, on average across all days




## Imputing missing values


Calculating the total number of rows with NAs

```{r, echo = TRUE}
TotalNA <- sum(!complete.cases(activityDF))
TotalNA 
```

The total number of rows with NAs is `r TotalNA` 


Creating a new dataset that is equal to the original dataset but with the missing values in 'steps' filled in with the means of the steps for the associated 5-minute intervals across all days


```{r, echo = TRUE}
newactivityDF <- activityDF

for (i in 1:nrow(newactivityDF)) {
   if (is.na(newactivityDF[i, 1])) {
      newactivityDF[i,1] <- averageStepsPerInterval$AverageSteps[averageStepsPerInterval$Interval == newactivityDF[i,3]]
      }
}

head(newactivityDF)
```

The total number of steps taken per day

```{r, echo = TRUE}
NewTotalStepsPerDay <- tapply(newactivityDF$steps, as.factor(newactivityDF$date), sum)
```



Histogram of total number of steps taken each day

```{r, echo = TRUE}
hist(NewTotalStepsPerDay, xlim = c(0,22000), ylim = c(0,20), breaks = 20, xlab = "Total Steps Per Day", main = "Histogram of Total Steps Per Day")
```


Mean of the total number of steps taken per day

```{r, echo = TRUE}
mean(NewTotalStepsPerDay)
```


Median of the total number of steps taken per day

```{r, echo = TRUE}
median(NewTotalStepsPerDay)
```


Imputing missing values with the means of the steps for 5-minute intervals hasn't had any effect on the mean and median values of the total steps per day. While the mean is identical, the median has not changed significantly.


## Are there differences in activity patterns between weekdays and weekends?


Creating a new factor variable in the dataset with two levels - "weekday" and "weekend"


```{r, echo = TRUE}
newactivityDF$Day <- ifelse(weekdays(newactivityDF$date) == "Saturday" | weekdays(newactivityDF$date) == "Sunday", "weekend", "weekday")
newactivityDF$Day <- as.factor(newactivityDF$Day)
```

Calculating the average number of steps for each 5-minute interval across all days grouped by weekends and weekdays


```{r, echo = TRUE}
newaverageStepsPerDay <- tapply(newactivityDF$steps, list(newactivityDF$interval, newactivityDF$Day), mean)
```

Converting newaverageStepsPerDay to a data frame


```{r, echo = TRUE}
newaverageStepsPerDay <- as.data.frame.table(newaverageStepsPerDay)
str(newaverageStepsPerDay)
```


Time series plot of the 5-minute interval and the average number of steps taken averaged across all weekdays or weekend days


```{r, echo = TRUE}
newaverageStepsPerDay$Var1 <- as.integer(levels(newaverageStepsPerDay$Var1))[newaverageStepsPerDay$Var1]
library(lattice)
xyplot(Freq~Var1|Var2, newaverageStepsPerDay, type = "l", layout = c(1,2), xlab = "interval", ylab = "Number of steps", xlim = c(0,2500))
```

